An Algorithm to Prune Integer Overflow Tests

### 2012-08-30 doc/Manuscript/CGO/paper.tex
Revised Raphael's section.

\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{mathpartir} % To use \inferrule
\usepackage{amssymb}    % To use \mathbb

@@

%What is our instrumentation library
We have implemented our instrumentation library as a LLVM transformation pass.
This means our library works on the LLVM IR\footnote{LLVM Intermediate Representation, available at \texttt{http://llvm.org/docs/LangRef.html}}
instead of working directly on the original source code. 
This fact brings the advantage of allowing us to rely on
other analysis, such as the range analysis, to check if it is safe to state that
the possible values of an instruction will always be within the range of
the declared type of the instruction. 
If we can state this, it means that it is impossible for an overflow to 
occur in the analyzed instruction, because the range analysis is a conservative analysis; 
in this case, we don't need to insert dynamic checks, and the instrumented program
will be smaller and run faster than it would run without this prunning.
%How our instrumentation works
Our instrumentation works by identifying the instructions that may 
lead to an overflow, and inserting assertions after that instructions. 
The LLVM IR has five instructions that may lead to an overflow: 
ADD, SUB, MUL, SHL, and TRUNC.

%ADD instrumentation
The ADD instruction performs the sum of two operands $op_1$ and $op_2$ and
stores the result in a variable $x$.
The semantics of an ADD instruction is $x = op_1 + op_2$. 
The goal of the instrumentation of the instruction $x$ is to ensure that after the
execution of the ADD instruction, $x$ stores the correct result of the sum,
without any data loss.
For this, we insert different checks for signed and unsigned instructions.
If we are dealing with an unsigned ADD instruction $x$, we insert the following check:
\code{if $x < op_1$ or $x < op_2$ then} handle overflow \code{else} continue to 
$x$'s next instruction. If, instead, we are instrumenting a signed instruction $x$, 
we will have the following assertion:
\code{if ($op_1 > 0$ and $op_2 > 0$ and $x < 0$) or ($op_1 < 0$ and $op_2 < 0$ and $x > 0$) then} 
handle overflow \code{else} continue to $x$'s next instruction.

%SUB instrumentation
The SUB instruction performs the subtraction of two operands $op_1$ and $op_2$ and
stores the result in a variable $x$.
The semantics of a SUB instruction is $x = op_1 - op_2$. 
Note that the SUB semantics is close to the ADD semantics. For this reason, the
instrumentation of the SUB instructions looks like the ADD instrumentation.
In this case, we insert different checks for signed and unsigned instructions, too.
If we are dealing with an unsigned SUB instruction $x$, we insert the following check:
\code{if $op_1 < op_2$ then} handle overflow \code{else} continue to 
$x$'s next instruction. Note that the assertion makes a single comparison. This is sound
because if $op_2$ is greater than $op_1$, the result would be negative. However, unsigned
integers can't represent negative numbers. In this case, we have a overflow.
The instrumentation of a signed SUB instruction $x$ will have the following assertion:
\code{if ($op_1 < 0$ and $op_2 > 0$ and $x > 0$) or ($op_1 > 0$ and $op_2 < 0$ and $x < 0$) then} 
handle overflow \code{else} continue to $x$'s next instruction.

%MUL instrumentation
The MUL instruction performs the multiplication of two operands $op_1$ and $op_2$ and
stores the result in a variable $x$.
The semantics of a MUL instruction is $x = op_1 \times{} op_2$. 
The dynamic check of the MUL instruction is the same for signed or unsigned
instructions, because it relies on DIV, the inverse operation. 
To verify if the overflow has occurred, we divide $x$ by $op_1$. If the result
is different of $op_2$, then we have an overflow to handle. 
In this case, we will have the following assertion: 
\code{if $op_1 \neq{} 0$ and $(x \div{} op_1) \neq{} op_2$ then} 
handle overflow \code{else} continue to $x$'s next instruction.
Note that if $op_1 = 0$, the result of the multiplication will never overflow,
because the result of the MUL will be $0$, too. In this case, if $op_1 = 0$,
we don't perform the DIV, to prevent a division by zero exception.
This technique of using the inverse operation to check for overflows doesn't work
for ADD and SUB instructions because the $n$-bits two's complement wrapping-arithmetics 
would give us false negatives in the assertions.

%SHL instrumentation
The SHL instruction performs the shift left operation. It receives a value $op_1$
and the number of bits $n$ the SHL will shift $op_1$ by; the operation's result
is stored in a variable $x$.
The semantics of a SHL instruction is $x = op_1 \times{} 2^n$. 
The SHL instructions result in overflows when the $op_1$ is positive and 
$x$ is negative or when $op_1$ is negative and $n \neq{} 0$, because 
the SHL in negative numbers would produce undefined results that are
as harmful as integer overflows. This assertion has the same semantics for
signed and unsigned instructions.
In this case, we will have the following check: 
\code{if ($op_1 > 0$ and $x < op_1$) or ($op_1 < 0$ and $n \neq{} 0$) then} 
handle overflow \code{else} continue to $x$'s next instruction.

%TRUNC instrumentation
The TRUNC instruction performs the truncation of the operand $op_1$
to a number $n$ of bits and stores the result in a variable $x$.
The semantics of a TRUNC instruction is $x = n$LessSignificantBits$(op_1)$.
The dynamic checking of a TRUNC instruction is done by expanding $x$ to
the datatype of $op_1$ and comparing the expanded $x$ with $op_1$. If they 
don't have the same value, some data has been lost by the TRUNC instruction.
In this case, we have the following check:
\code{if $cast(x, getDataType(op_1)) \neq{} op_1$ then} 
handle overflow \code{else} continue to $x$'s next instruction.
Note that the instrumentation of the TRUNC instructions is a very fragile point, 
because it would catch any truncation with loss of precision, even if it is a
benign one.
To avoid trouble and make it easier to our users, the instrumentation library 
can receive an argument to let the user decide whether to insert or not
instrumentation in the TRUNC instructions.

%Overview about the instrumentation technique
Our instrumentation library works by adding new instructions to
the target programs. The instrumentation of all the instructions works
in the same way: first, the pass inserts instructions to gather the
information needed to do the assertion.
Then, the pass inserts compare instructions to generate a boolean value
that indicates if there is an overflow or not. 
After this, the pass inserts a conditional branch: 
the false clause of the branch, where the control goes when there is no 
overflow, leads to the first instruction after the instruction
which has received the instrumentation.
the true clause of the branch leads to a block where we do the
overflow handling.

>>>>

\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{mathpartir} % To use \inferrule
\usepackage{amssymb}    % To use \mathbb
\usepackage{skak}

@@

%What is our instrumentation library
We have implemented our instrumentation library as a LLVM transformation pass;
thus, we work at the level of the compiler's intermediate
representation~\footnote{\texttt{http://llvm.org/docs/LangRef.html}}.
This is in contrast to previous work, which either transforms the
source code~\cite{Dietz12}, or the machine dependent code~\cite{Brumley07}.
We work at this level because we want to couple our library with static
analyses, such as the algorithm that we describe in Section~\ref{sec:range}.
% TODO: add text about the interaction between static an dynamic analysis at
% the end of this section.
Our instrumentation works by identifying the instructions that may 
lead to an overflow, and inserting assertions after those instructions. 
The LLVM IR has five instructions that may lead to an overflow:
\textsc{Add}, \textsc{Sub}, \textsc{Mul},  \textsc{Trunc} and \textsc{Shl}
(left shift).
Figure~\ref{fig:instrumentation} shows the conditions that we use to
detect overflows.

\begin{figure}[t!]
\begin{center}
\begin{tabular}{ll}
$x = o_1 \ +_s \ o_2$ & $(o_1 > 0 \wedge o_2 > 0 \wedge x < 0) \ \ \vee$ \\
                  & $(o_1 < 0 \wedge o_2 < 0 \wedge x > 0)$ \\ \\
$x = o_1 \ +_u \ o_2$ & $x < o_1 \vee x < o_2$ \\ \\
$x = o_1 \ -_s \ o_2$ & $(o_1 < 0 \vee o_2 > 0 \vee x > 0) \vee$ \\
                  & $(o_1 > 0 \vee o_2 < 0 \vee x < 0)$ \\ \\
$x = o_1 \ -_u \ o_2$ & $o_1 < o_2$ \\ \\
$x = o_1 \ \times_{u/s} \ o_2$ & $x \neq 0 \Rightarrow x \div o_1 \neq o_2$ \\ \\
$x = o_1 \ \qside \ n$ & $(o_1 > 0 \wedge x < o_1) \vee (o_1 < 0 \wedge n \neq 0)$ \\ \\
$x = \ \downarrow_n \ o_1$ & $\mbox{cast}(x, \mbox{type}(o_1)) \neq o_1$ \\ \\
$x = \ (\mbox{type}_n) o_1$ & $\mbox{cast}(x, \mbox{type}(o_1)) \neq o_1$ \\
\end{tabular}
\end{center}
\caption{\label{fig:instrumentation}Overflow checks. We use $\downarrow_n$ for
the operation that truncates to $n$ bits, and we use $(\mbox{type}_n)$ for a
bit-cast to $n$ bits. Both operations have the same semantics in our setting.
The subscript $s$ indicates a signed instruction; the subscript $u$ indicate
an unsigned operation.}
\end{figure}

%SUB instrumentation
The instrumentation that we insert is mostly straightforward.
We discuss in the rest of this section a few cases.
If we are dealing with an unsigned \textsc{Sub} instruction, e.g,
$x = o_1 \ -_u \ o_2$, then a single check is enough to detect the bad
behavior: $o_1 < o_2$.
This is sound because if $o_2$ is greater than $o_1$, then we assume that it is
a bug to try to represent a negative number in unsigned arithmetics.
Regarding multiplication, e.g., $x = o_1 \times o_2$, if $o_1 = 0$, then this
operation can never cause an overflow.
This test is necessary to us, because we check integer overflows in
multiplication via the inverse operation of integer division.
Thus, the test prevents a division by zero from happening.
The semantics of a \textsc{Trunc} instruction $x = \ \downarrow_n \ o_1$ is
to assign to $x$ the $n$ least significant bits of $o_1$.
The dynamic checking in this case consists in expanding $x$ to
the datatype of $o_1$ and comparing the expanded value with $o_1$.
The LLVM IR provides instructions to perform these type expansions.
Note that our instrumentation catches any truncation with loss of precision, even
if it is a benign one.
However, we give users the possibility of disabling checks on truncation.

\paragraph{Practical Considerations: }
%Overview about the instrumentation technique
Our instrumentation library works by adding new instructions to
the target programs.
The instrumentation of all the instructions works
in the same way: first, the pass inserts instructions to gather the
information needed to do the assertion.
Then, the pass inserts compare instructions to generate a boolean value
that indicates if there is an overflow or not. 
After this, the pass inserts a conditional branch: 
the false clause of the branch, where the control goes when there is no 
overflow, leads to the first instruction after the instruction
which has received the instrumentation.
the true clause of the branch leads to a block where we do the
overflow handling.
